On the formation of H$\alpha$ line emission around classical T Tauri stars

### 2006-02-02 TTau_Halpha/submit/kurosawa.tex
Right after the submission.
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@653 25464f21-6099-402c-b30a-640f1d40d38b

Despite these successes, the overwhelming observational evidence for
outflows in CTTS suggests that the MA model is only one component
of a complex circumstellar environment. Clearly, one must include the
contribution of any wind/jet flow if one wishes to both accurately
predict the mass-accretion rate and also determine the mass-loss rate
of a CTTS via emission profile modelling. The first attempt in this
direction was made by \citet{alencar:2005} who demonstrated that the
observed H$\alpha$, H$\beta$ and Na~D lines of RW Aur are better
reproduced by the radiative transfer model which included a collimated
disc-wind arising from near the inner edge of the accretion disc.

@@

Assuming mass-flux conservation and using the velocity field defined
above, the disc wind density as a function of $w$ and $l$ can be
written as
\begin{equation}
  \rho\left(w_{i},l\right) =
  \frac{\dot{m}\left(w_{i}\right)}{v_{p}\left(w_{i},l\right)\,\left|\cos\delta\right|}\,\left\{ 
  \frac{d}{D\left(w_{i},l\right)\,\cos\delta}\right\} ^{2}
  \label{eq:discwind-density}
\end{equation} 
where $D$ and $\delta$ are the distance from the source point ($S$) to
a point along the stream line and the angle between the stream line
and the disc normal, respectively. Fig.~\ref{fig:discwind_vr_rho}
shows the density and the velocity components along the mid stream
line (passing through
$w_{i}=\left(R_{\mathrm{wi}}+R_{\mathrm{wo}}\right)/2$ on the disc
plane ($z=0$) for different values of the wind acceleration parameter
$\beta$.

@@

  \caption{The dependency of the H$\alpha$ profile on inclination ($i$).  The
    profiles are computed with a magnetospheric accretion flow only using
    the reference parameters given in Table~\ref{tab:std_parameters} and
    $T_{\mathrm{max}}=7500~\Kelvin$. The solid, dotted, and dashed lines
    are for $i=10^{\circ}$, $55^{\circ}$, and $80^{\circ}$
    respectively. As the inclination becomes larger, the peak flux and the
    equivalent width of the line becomes smaller.  }

@@

The intensity of the emission component increases as
$\beta$ increases for a fixed $T_{\mathrm{wind}}$. This is mainly
because the density near the photosphere increases as the value of
$\beta$ increases (Fig.~\ref{fig:discwind_vr_rho}); hence, the
emissivity of increases. Most of the profiles show a red-ward asymmetry, mainly
due to the presence of the classical P-Cygni absorption. However we find
that the disc-wind models with the fastest acceleration ($\beta=0.2$ and $0.5$)
and viewed at low inclinations, show a strong blue asymmetry due to
occultation of the receding half of the outflow by the optically thick disc.

@@

As in the disc-wind only models
(Fig.~\ref{fig:atlas_discwind_only}), the location of the absorption
component moves toward the line centre as the value of $\beta$
increases.  The figure shows that in the models with smaller $\beta$
and $T_{\mathrm{wind}}$, the line emission from the magnetosphere
dominates, while this situation reverses for the models with larger
$\beta$ and $T_{\mathrm{wind}}$. The model with
$\left(T_{\mathrm{wind}},\,\beta\right)=\left(9000\,\mathrm{K},\,2.0\right)$
results in an H$\alpha$ profile that is far too strong to be
compatible with observations (\citealt{reipurth:1996} and
\citealt{alencar:2000}); however, the lower mass-loss rate models
(while keeping $\mu$ constant) produce profiles with the line
strengths comparable to observations (c.f.~appendix~\ref{sec:appendix}).
Although the models in Fig.~\ref{fig:atlas_hybrid} are computed with
limited ranges of $T_{\mathrm{wind}}$, $\beta$ and a fixed $i$, the
resulting profiles exhibit a wide variety of line profile shapes, many
of which are similar to the types of H$\alpha$ profiles seen in the
observations (c.f.~\citealt{reipurth:1996}).

@@

By inspecting the model H$\alpha$ images computed
(Fig.~\ref{fig:discwind_hybrid_ha_image}), we have 
found that the extent of the H$\alpha$ emission region is relatively
small (i.e.~$h\sim 0.1\,\mathrm{au}$ or $10\, R_{*}$ where $h$ is the
vertical distance from the disc) when compared to the
spectro-astrometric observations of \citet{takami:2003} 
for RU~Lup and CS Cha which show 1--5~au scale outflows%
\footnote{Their observations also show some objects
(e.g~Z CMa and AS~353A) displaying the outflows in larger scale
($>50\,\mathrm{au}$); however, this could be formed in shocks rather
than MHD-wave heating (e.g. \citealt{hartmann:1982}) or X-ray heating
(e.g.~\citealt{shang:2002}). %
}.  In our model, the wind emission flux drops by $\sim10^{4}$ by
the time the wind reaches $h \approx 0.1\,\mathrm{au}$. The extent of the line
emission region becomes slightly larger ($\sim0.4$~au) if a slower
wind acceleration rate (e,g,~$\beta=4$) is used. The discrepancy with
the observations is perhaps caused by the lack of the material
connecting the \emph{disc wind} to a collimated \emph{jet} in our
model, which may be excited by alternative heating mechanism
(e.g.~shocks) thus producing observable H$\alpha$ emission at large
distances from the central object.

@@

\citet{reipurth:1996} proposed a two-dimensional classification
of H$\alpha$ emission profiles of T~Tauri stars and Herbig Ae/Be
stars. Their classification scheme contains four classes (I, II, III
and IV) differentiated by the ratio of the secondary-to-primary emission
components in the profiles. Each class is divided into two sub-classes
(B and R) which depends whether the secondary peak is on the
blue or red side of the primary peak. Readers are referred to Fig.~4 of their paper.
Fig.~\ref{fig:atlas_reipurth} shows  sample model profiles
which are classified according to the definition of \citet{reipurth:1996}.
The combination of the disc wind, magnetospheric accretion flow, and
the accretion disc can reproduce all the classes of the profiles
seen in observations. The corresponding parameters of the
disc-wind-magnetosphere hybrid model used to reproduce the profiles
in the figure are summarised in Table~\ref{tab:classification} along
with brief comments on possible physical conditions which lead to
the profiles in each class. Below we discuss the Reipurth scheme in the
context of our models parameters. Readers are referred to
appendix~\ref{sec:appendix} for the complete model profiles used in 
this discussion.

@@

  \caption{An H$\alpha$ atlas of the disc-wind-magnetosphere hybrid
  model with $\dot{M}_{\mathrm{acc}}
  =10^{-7}\,\mathrm{M_{\sun}\,yr{-1}}$ and
  $\dot{M}_{\mathrm{wind}}=10^{-6}\,\mathrm{M_{\sun}\,yr{-1}}$.
  Profiles are presented for inclinations of $10^\circ$ (top figure),
  $55^\circ$ (middle figure), and $80^\circ$ (bottom figure).  For
  each inclination, 12 profiles are given, for three different values
  of the wind temperature, and four different values of the wind
  acceleration parameter $\beta$. The contribution of the
  magnetosphere can be judged from the profiles computed with the
  magnetosphere only (dotted lines).}

@@

  \caption{As for Fig.~\ref{fig:atlas01} except
  $\dot{M}_{\mathrm{acc}} =10^{-8}\,\mathrm{M_{\sun}\,yr{-1}}$ and
  $\dot{M}_{\mathrm{wind}}=10^{-7}\,\mathrm{M_{\sun}\,yr{-1}}$. }

@@

  \caption{As for Fig.~\ref{fig:atlas01} except
    $\dot{M}_{\mathrm{acc}} =10^{-9}\,\mathrm{M_{\sun}\,yr{-1}}$ and
  $\dot{M}_{\mathrm{wind}}=10^{-10}\,\mathrm{M_{\sun}\,yr{-1}}$. }

>>>>

Despite these successes, the overwhelming observational evidence for
outflows in CTTS suggests that the MA model is only one component
of a complex circumstellar environment. Clearly, one must include the
contribution of any wind/jet flow if one wishes to both accurately
predict the mass-accretion rate and also determine the mass-loss rate
of a CTTS via emission profile modelling. The first attempt in this
direction was made by \citet{alencar:2005} who demonstrated that the
observed H$\alpha$, H$\beta$ and Na~D lines of RW~Aur are better
reproduced by the radiative transfer model which included a collimated
disc-wind arising from near the inner edge of the accretion disc.

@@

Assuming mass-flux conservation and using the velocity field defined
above, the disc wind density as a function of $w_{i}$ and $l$ can be
written as
\begin{equation}
  \rho\left(w_{i},l\right) =
  \frac{\dot{m}\left(w_{i}\right)}{v_{p}\left(w_{i},l\right)\,\left|\cos\delta\right|}\,\left\{ 
  \frac{d}{D\left(w_{i},l\right)\,\cos\delta}\right\} ^{2}
  \label{eq:discwind-density}
\end{equation} 
where $D$ and $\delta$ are the distance from the source point ($S$) to
a point along the stream line and the angle between the stream line
and the disc normal, respectively. Fig.~\ref{fig:discwind_vr_rho}
shows the density and the velocity components along the mid stream
line (passing through
$w_{i}=\left(R_{\mathrm{wi}}+R_{\mathrm{wo}}\right)/2$ on the disc
plane ($z=0$) for different values of the wind acceleration parameter
$\beta$.

@@

  \caption{The dependency of the H$\alpha$ profile on inclination ($i$).  The
    profiles are computed with a magnetospheric accretion flow only using
    the reference parameters given in Table~\ref{tab:std_parameters} and
    $T_{\mathrm{max}}=7500~\Kelvin$. The solid, dotted, and dashed lines
    are for $i=10^{\circ}$, $55^{\circ}$, and $80^{\circ}$
    respectively. As the inclination becomes larger, the peak flux and the
    equivalent width of the line become smaller.  }

@@

The intensity of the emission component increases as
$\beta$ increases for a fixed $T_{\mathrm{wind}}$. This is mainly
because the density near the photosphere increases as the value of
$\beta$ increases (Fig.~\ref{fig:discwind_vr_rho}); hence, the
emissivity increases. Most of the profiles show a red-ward asymmetry, mainly
due to the presence of the classical P-Cygni absorption. However we find
that the disc-wind models with the fastest acceleration ($\beta=0.2$ and $0.5$)
and viewed at low inclinations, show a strong blue asymmetry due to
occultation of the receding half of the outflow by the optically thick disc.

@@

As in the disc-wind only models
(Fig.~\ref{fig:atlas_discwind_only}), the location of the absorption
component moves toward the line centre as the value of $\beta$
increases.  The figure shows that in the models with smaller $\beta$
and $T_{\mathrm{wind}}$, the line emission from the magnetosphere
dominates, while this situation reverses for the models with larger
$\beta$ and $T_{\mathrm{wind}}$. The model with
$\left(T_{\mathrm{wind}},\,\beta\right)=\left(9000\,\mathrm{K},\,2.0\right)$
results in an H$\alpha$ profile that is far too strong to be
compatible with observations (\citealt{reipurth:1996}; 
\citealt{alencar:2000}); however, the lower mass-loss rate models
(while keeping $\mu$ constant) produce profiles with the line
strengths comparable to observations (c.f.~Appendix~\ref{sec:appendix}).
Although the models in Fig.~\ref{fig:atlas_hybrid} are computed with
limited ranges of $T_{\mathrm{wind}}$, $\beta$ and a fixed $i$, the
resulting profiles exhibit a wide variety of line profile shapes, many
of which are similar to the types of H$\alpha$ profiles seen in the
observations (c.f.~\citealt{reipurth:1996}).

@@

By inspecting the model H$\alpha$ images computed
(Fig.~\ref{fig:discwind_hybrid_ha_image}), we have 
found that the extent of the H$\alpha$ emission region is relatively
small (i.e.~$h\sim 0.1\,\mathrm{au}$ or $10\, R_{*}$ where $h$ is the
vertical distance from the disc) when compared to the
spectro-astrometric observations of \citet{takami:2003} 
for RU~Lup and CS Cha which show 1--5~au scale outflows%
\footnote{Their observations also show some objects
(e.g~Z CMa and AS~353A) displaying the outflows in larger scale
($>50\,\mathrm{au}$); however, this could be formed in shocks rather
than MHD-wave heating (e.g. \citealt{hartmann:1982}) or X-ray heating
(e.g.~\citealt{shang:2002}). %
}.  In our model, the wind emission flux drops by $\sim10^{4}$ by
the time the wind reaches $h \approx 0.1\,\mathrm{au}$. The extent of the line
emission region becomes slightly larger ($\sim0.4$~au) if a slower
wind acceleration rate (e,g,~$\beta=4$) is used. The discrepancy
between the observations and the model is perhaps caused by the lack
of the material connecting the \emph{disc wind} to a collimated
\emph{jet} in our model, which may be excited by alternative heating
mechanism (e.g.~shocks) thus producing observable H$\alpha$ emission
at large distances from the central object.

@@

\citet{reipurth:1996} proposed a two-dimensional classification
of H$\alpha$ emission profiles of T~Tauri stars and Herbig Ae/Be
stars. Their classification scheme contains four classes (I, II, III
and IV) differentiated by the ratio of the secondary-to-primary emission
components in the profiles. Each class is divided into two sub-classes
(B and R) which depends whether the secondary peak is on the
blue or red side of the primary peak. Readers are referred to Fig.~4 of their paper.
Fig.~\ref{fig:atlas_reipurth} shows  sample model profiles
which are classified according to the definition of \citet{reipurth:1996}.
The combination of the disc wind, magnetospheric accretion flow, and
the accretion disc can reproduce all the classes of the profiles
seen in observations. The corresponding parameters of the
disc-wind-magnetosphere hybrid model used to reproduce the profiles
in the figure are summarised in Table~\ref{tab:classification} along
with brief comments on possible physical conditions which lead to
the profiles in each class. Below we discuss the Reipurth scheme in the
context of our models parameters. Readers are referred to
Appendix~\ref{sec:appendix} for the complete model profiles used in 
this discussion.

@@

  \caption{An H$\alpha$ atlas of the disc-wind-magnetosphere hybrid
  model with $\dot{M}_{\mathrm{acc}}
  =10^{-7}\,\mathrm{M_{\sun}\,yr^{-1}}$ and
  $\dot{M}_{\mathrm{wind}}=10^{-6}\,\mathrm{M_{\sun}\,yr^{-1}}$.
  Profiles are presented for inclinations of $10^\circ$ (top figure),
  $55^\circ$ (middle figure), and $80^\circ$ (bottom figure).  For
  each inclination, 12 profiles are given, for three different values
  of the wind temperature, and four different values of the wind
  acceleration parameter $\beta$. The contribution of the
  magnetosphere can be judged from the profiles computed with the
  magnetosphere only (dotted lines).}

@@

  \caption{As for Fig.~\ref{fig:atlas01} except
  $\dot{M}_{\mathrm{acc}} =10^{-8}\,\mathrm{M_{\sun}\,yr^{-1}}$ and
  $\dot{M}_{\mathrm{wind}}=10^{-7}\,\mathrm{M_{\sun}\,yr^{-1}}$. }

@@

  \caption{As for Fig.~\ref{fig:atlas01} except
    $\dot{M}_{\mathrm{acc}} =10^{-9}\,\mathrm{M_{\sun}\,yr^{-1}}$ and
  $\dot{M}_{\mathrm{wind}}=10^{-10}\,\mathrm{M_{\sun}\,yr^{-1}}$. }
